Type theory as a quotient inductive-recursive type

### 2025-01-02 tex/main.tex
[ tex ]

\begin{abstract}
  Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are painstakingly littered with explicit coercions, i.e.\ \transp's, along equality constructors.
  This mess has been dubbed the \emph{transport hell} by the community in private but typically suppressed in public presentation for clarity, ignoring the elephant in the room.
  In this paper, we aim for shortening the gap between formal and informal definitions of type theory in type theory using quotient inductive-inductive-recursive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as QIIRTs with meta-theoretic properties proved and compare our definitions with quotient inductive-inductive definitions.
\end{abstract}

@@

\section{Metatheory and formalisation}
\cite{Pujet2022,Pujet2024,Pujet2022a}
Points to include:
\begin{enumerate}
  \item dependent identity type (heterogeneous equality in \Agda)
  \item 
\end{enumerate}

@@

\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}} \\
  \emptyctx\eta   & : \implicit{\Gamma, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t \\
  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t} \\
\end{alignat*}

\begin{example}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$. 
  Intuitively, this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ applied by the substitution $\sigma$. 
  That is, the following equality apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}

  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, more verbosely,
  \[
    \transfib{\Tm\,\Gamma}{p}{\pi_2\,(\sigma ; \tau)} =^{\Tm\,\Gamma\,([\sigma][\pi_1\,\tau]\,A)}  [\sigma] (\pi_2\,\tau)
  \]
  where $p$ is a term of the identity $[\pi_1(\sigma; \tau)]\,A = [\sigma]\,[\pi_1\,\tau]A$.
  Now that we have a transported term we have to carry out our proof with coercion rules explicitly, and moreover in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ and $\pi_2\beta$ which introduced another transported terms.
  It turns out that we have to `eliminate' these $\transp$'s to derive the right-hand side of~\eqref{eq:pi2-comp-proof}, making the actual proof barely readable.
  \todo{write down the actual proof}
\end{example}
\begin{alignat*}{3}
[ \idS ]_{\Ty}\,A             & \jdeq A \\
[ \sigma ; \tau ]_{\Ty}\,A    & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
[ \pi_1(\sigma, t) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty\,A \\
[ \sigma ]_{\Ty}\,U           & \jdeq U
\end{alignat*}

>>>>

\begin{abstract}
  Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are painstakingly littered with explicit coercions, i.e.\ \transp's, along equality constructors.
  This mess has been dubbed the \emph{transport hell} by the community in private but typically suppressed in presentation for clarity, ignoring the elephant in the room.
  In this paper, we aim for shortening the gap between formal and informal definitions of type theory in type theory using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as quotient inductive-inductive-recursive types with meta-theoretic properties proved and compare our definitions with quotient inductive-inductive definitions.
\end{abstract}

@@

\section{Metatheory and formalisation}
\cite{Pujet2022,Pujet2024,Pujet2022a}
Points to include:
\begin{enumerate}
  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality in \Agda $x \simeq y$ for $x : A$ and $y : B$
  \item 
\end{enumerate}

@@

\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}} \\
  \emptyctx\eta   & : \implicit{\Gamma, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t \\
  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t} \\
\end{alignat*}

\begin{example}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following equality apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, equivalently
  \[
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \]
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to eliminate that $\transp$ to derive the right hand side.

  To better illustrate the annoyance, we provide a complete proof.
  First, we claim each of the following steps holds:
  \begin{alignat}{3}
         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \label{eq:pi2-proof-1} \\
    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \nonumber \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \nonumber \\
    = {} & (\sub{\sigma;\pi_1\tau}{A} &&, [\sigma] (\pi_2\,\tau)) \nonumber
  \end{alignat}
  where $p \judgeq [;]_{\Ty}^{-1} : \sub{\sigma}{\sub{\pi_1\,\tau}{A}} \to \sub{\sigma; \pi_1 \tau}{A}$ and $t \judgeq\transfib{\Tm\,\Gamma}{p}{[\sigma]\,(\pi_2\,\tau)}$.
\end{example}

\begin{alignat*}{3}
[ \idS ]_{\Ty}\,A             & \jdeq A \\
[ \sigma ; \tau ]_{\Ty}\,A    & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
[ \pi_1(\sigma, t) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty\,A \\
[ \sigma ]_{\Ty}\,U           & \jdeq U
\end{alignat*}
